---
title: "Module 3 - BLAST (megablast) and organizing output"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This module starts from the trimmed `.fasta` exported in **Module 2** and performs a **megablast** search against NCBI, then exports a table of BLAST hits with **taxonomy (organism + taxid)**.

## 1. Load packages

```{r, eval=TRUE, message=FALSE}
library(tidyverse)
```

The BLAST function uses:

- `Biostrings` to read FASTA
- `httr` to talk to NCBI BLAST URL API
- `rentrez` to retrieve taxonomy metadata from NCBI (optional but recommended)

```{r, eval=FALSE}
# Install if needed
install.packages("httr")
install.packages("rentrez")
install.packages("xml2")   # optional (for lineage best-effort)
BiocManager::install("Biostrings")
```

## 2. Run megablast from R

This pipeline uses the script `R_script/5_megablast_fasta.R`.

```{r, eval=FALSE}
setwd("D:/MSL Miscellaneous/20240408 BGI Pipeline Github")
source("R_script/5_megablast_fasta.R")
```

### Input FASTA

Use the example FASTA shipped with this repo:

```{r, eval=FALSE}
fasta_path <- "example_data/16S_trimmed_seq.fasta"
```

Or use your own FASTA produced in Module 2 (replace the path with your file):

```{r, eval=FALSE}
fasta_path <- "path/to/your/trimmed_sequences.fasta"
```

### Output files

There are **two outputs**:

- A **per-query temporary folder** containing one CSV per query sequence
- A **final merged CSV** containing all queries

```{r, eval=FALSE}
results <- run_megablast(
  fasta_path = fasta_path,
  output_path = "example_data/16S_megablast_results_full.csv",
  # NCBI requests an email/tool name for automated usage
  email = "your_email@example.com",
  tool = "seaslug_barcoding",
  hitlist_size = 50,
  include_taxonomy = TRUE,
  # temp files will be written here (default is dirname(output_path)/megablast_temp)
  temp_dir = "example_data/megablast_temp"
)
```

The returned `results` object is a data frame with (at minimum):

- `query_id`
- `subject_id` (accession.version)
- alignment statistics (`pident`, `length`, `evalue`, `bitscore`, etc.)
- taxonomy columns (when `include_taxonomy = TRUE`): `organism`, `taxid` (and `lineage` when available)

## 3. Notes on NCBI usage and runtime

- **NCBI rate limits**: the script polls BLAST jobs at most about once per minute per request ID (RID), and taxonomy queries are throttled with `Sys.sleep()` to be polite to NCBI servers.
- **Long runs**: BLAST for many sequences can take a long time. Writing per-query results into `temp_dir` ensures you keep partial results even if the run is interrupted.
- **Taxonomy completeness**: taxonomy is derived from NCBI `nuccore` summaries. Some accessions can fail intermittently; the script performs a **second-pass fallback** using `entrez_search()` (accession → UID) then `entrez_summary()` until no new matches (bounded rounds).

## 4. Troubleshooting

### BLAST polling looks stuck

If you see a RID and want to check it manually:

```{r, eval=FALSE}
# Replace RID with your value
# https://blast.ncbi.nlm.nih.gov/Blast.cgi?CMD=Get&RID=YOUR_RID
```

### Parse failures

If you suspect NCBI returned an unexpected format, re-run with:

```{r, eval=FALSE}
results <- run_megablast(
  fasta_path = fasta_path,
  output_path = "example_data/16S_megablast_results_full.csv",
  email = "your_email@example.com",
  debug_save_raw = TRUE
)
```

This writes the raw response for debugging when parsing fails.

